---
title: "molar"
output: html_document
---

notes 

- M3 missing when m2/m1 ~ 0.4 
- M2/m1 increases as inhibition decreases
- lack of m3 corr with shorter mandible (fabio)

- is there any reason for a trend at all? (why/why not)

- seems to match the ICM trend, I guess? How could you falsifey this? what would be the difference betwee it being a neutral relationship vs heavily selected 

- old world monkeys have outlier molars 

```{r}
# make tree data 
make.treedata <- function(tree, data, name_column="detect", as.is=FALSE) {
  if(!inherits(tree, "phylo")) stop("tree must be of class 'phylo'")
  if(is.vector(data)){
    data <- as.matrix(data)
    colnames(data) <- "trait"
  }
  if(is.null(colnames(data))){
    colnames(data) <- paste("trait", 1:ncol(data), sep="")
  }
  coln <- colnames(data)
  if(name_column=="detect"){
    if(is.null(rownames(data))){
      tmp.df <- data.frame(data)
      offset <- 0
    } else {
      tmp.df <- data.frame(rownames(data), data)
      offset <- 1
    }
    matches <- sapply(tmp.df, function(x) sum(x %in% tree$tip.label))
    if(all(matches==0)) stop("No matching names found between data and tree")
    name_column <- which(matches==max(matches))-offset
  } else{
    if(is.character(name_column)){
      name_column <- which(name_column==coln)[1]
    }
  }
  dat <- as_tibble(as.data.frame(lapply(1:ncol(data), function(x)
    type.convert(apply(data[,x, drop=FALSE], 1, as.character), as.is=as.is))))
  colnames(dat) <- coln
  if(name_column==0){
    clnm <- colnames(dat)
    dat <- dat[,clnm, drop=FALSE]
    dat.label <- as.character(rownames(data))
  } else {
    if(is.numeric(name_column)){
      clnm <- (1:ncol(data))[-name_column]
    } else {
      clnm <- colnames(dat)[-which(colnames(dat)==name_column)]
    }
    dat <- dat[, clnm, drop=FALSE]
    dat.label <- as.character(as.data.frame(data)[[name_column]])
  }
  data_not_tree <- setdiff(dat.label, tree$tip.label)
  tree_not_data <- setdiff(tree$tip.label, dat.label)
  phy <- drop.tip(tree, tree_not_data)
  dat <- filter(dat, dat.label %in% phy$tip.label)
  dat.label <- dat.label[dat.label %in% phy$tip.label]
  if(any(duplicated(dat.label))){
    warning("Duplicated data in dataset, selecting first unique entry for each species")
    dat <- filter(dat, !duplicated(dat.label))
    dat.label <- dat.label[!duplicated(dat.label)]
  }
  ...my.order... <- match(dat.label, phy$tip.label)
  dat <- arrange(dat, ...my.order...)
  td <- list(phy=phy, dat=dat)
  class(td) <- c("treedata", "list")
  attributes(td)$tip.label <- phy$tip.label
  #attributes(td$dat)$row.names <- phy$tip.label
  attributes(td)$dropped <- list(dropped_from_tree=data_not_tree,dropped_from_data=tree_not_data)
  #Removing rownames for future versions of dplyr
  #rownames(td$dat) <- attributes(td)$tip.label
  return(td)
}
```

```{r}
# Load required libraries
library(ape)
library(dplyr)

# Read in data
prim_tree <- read.nexus("../data/prim_10k.nex")
dat <- read.csv("../data/Plavcan_Lower Molar_All.csv")
new_dat <- dat
new_dat[is.na(new_dat)] <- 0  # Replace NA values with 0

# Calculate presence/absence of the third molar
p_dat <- dat %>%
  mutate(pres.m3 = ifelse(is.na(m3.md), 0, 1)) %>%
  select(Species, pres.m3) %>%
  rename(taxa = Species)

# Fixing a known subspecies naming issue
p_dat$taxa[p_dat$taxa == "Semnopithecus_entellus_lianna"] <- "Semnopithecus_entellus"

# Calculate the ratio of presence for each species
counts <- p_dat %>%
  group_by(taxa) %>%
  summarise(ratio = mean(pres.m3))

# Transform the ratio data into discrete categories
d_dat <- counts %>%
  mutate(pres = ifelse(ratio < 1, "0", "1"))

# Prepare the threshold data
thresh_dat <- d_dat %>%
  mutate(pres = ifelse(pres == "1", "3&4", "1&2")) %>%
  select(taxa, pres)

# Output the data frames for inspection
print(d_dat)
print(thresh_dat)
```
## subspecies pres/absence 

```{r}
library(dplyr)

# Assuming p_dat already has the taxa and pres.m3 columns
# Extract genus and species from taxa, ignoring subspecies
p_dat$genus_species <- sapply(strsplit(as.character(p_dat$taxa), "_"), function(x) paste(x[1:min(2,length(x))], collapse = "_"))

# Summarize at the species level, considering unique taxa
species_summary <- p_dat %>%
  group_by(genus_species) %>%
  summarise(
    distinct_taxa_count = n_distinct(taxa),
    presence_count = sum(pres.m3, na.rm = TRUE),
    absence_count = sum(1 - pres.m3, na.rm = TRUE),
    presence_ratio = mean(pres.m3, na.rm = TRUE)
  )

# View the summarized species data
print(species_summary)
```
```{r}
library(dplyr)

# Assuming species_summary is already created with genus_species, presence_ratio, etc.
# Define a threshold for presence/absence
threshold <- 1 # Adjust this threshold as per your analysis requirements

# Discretize the presence/absence data
species_discrete <- species_summary %>%
  mutate(pres_absence = ifelse(presence_ratio >= threshold, 1, 0)) %>%
  select(genus_species, pres_absence)

# Rename columns for consistency
colnames(species_discrete) <- c("taxa", "pres_absence")

# View the new dataframe
print(species_discrete)


# making thresh data
t_dat <- species_discrete

t_dat$pres_absence <- case_when(
   t_dat$pres_absence == 0 ~ "1&2",
  t_dat$pres_absence == 1 ~ "3&4"
)
```



## species list cleaning 

```{r}
temp_dat <- d_dat
# removing function
sp <- sort(sp)

drop_subspecies <- function(x){
y <- strsplit(x, "_")[[1]]
z <- paste(y[1], y[2], sep="_")
return(z)
}

new <- list()

# update this at a later time 
some_trim <- c(5,7,8,9,25,11,15,25,26,27,30)
sp_drop <- sp[some_trim]

    for (k in 1:length(some_trim)) {
      new[k] <- drop_subspecies(sp_drop[k])
    }


# trying to replace subspecies with general sp.
place <- which(d_dat$taxa %in% sp_drop) 
new <- unlist(new)
new <- new[-c(4,8)]
d_dat$taxa[place] <- new


new2 <- new 
place2 <- which(thresh_dat$taxa %in% sp_drop) 
new2 <- unlist(new2)
new2 <- new2[-c(4,8)]

thresh_dat$taxa[place2] <- new2
# removing duplicates 
d_dat <- d_dat[!duplicated(d_dat),]
thresh_dat <- thresh_dat[!duplicated(thresh_dat),]
```

```{r}
drop_subspecies <- function(x){
  parts <- strsplit(x, "_")[[1]]
  if (length(parts) > 2) {
    # If there are more than two parts, assume it's a subspecies and remove it
    return(paste(parts[1], parts[2], sep = "_"))
  } else {
    # If there are only two parts, return the name as it is
    return(x)
  }
}
results <- lapply(sp, drop_subspecies)

new_names <- sapply(results, function(x) x$new_name)
is_subspecies <- sapply(results, function(x) x$is_subspecies)
some_trim <- which(is_subspecies)


```

# analyses 




```{r}
# matching tree & analysis 
library(treeplyr)
library(corHMM)
td <- make.treedata(prim_tree, species_discrete)
td_thresh <- make.treedata(prim_tree, t_dat)

head(td$dat)
corr_d <- corHMM::corHMM(phy = td$phy, data = d_dat, rate.cat = 2)

corr_ER <- corHMM::corHMM(phy = td$phy, data = d_dat, rate.cat = 2, model = "ER")

corr_two <- corHMM(phy = td$phy, data = d_dat, rate.cat = 4)

## thresh model
three <- matrix( c(NA,1,NA,1,NA,1,NA,1,NA), nrow = 3, ncol= 3)
rownames(three) <- c("1","2","3")
colnames(three) <- c("1","2","3")
four <- matrix(c(NA,1,NA,NA,NA,NA,1,NA,NA,1,NA,NA,NA,NA,1,NA), nrow=4, ncol=4)
rownames(four) <- c("1","2","3","4")
colnames(four) <- c("1","2","3","4")

#corr_thresh <- corHMM(phy = td$phy, d_dat, rate.mat = four,rate.cat = 2)
 
thresh_dat <- case_when(
  d_dat$pres == 1 ~ "3&4",
  d_dat$pres == 0 ~ "1&2"
)

# I Guess I need to fix the roots?
root.obs <- function(prop, bins, cutoff=3.1){
  qq <- qnorm(prop, 0, 1)
  x <- pnorm(seq(-1*cutoff, cutoff, length.out=bins-1), qq,1)
  x <- c(x[2:(length(x)-1)],1) - c(0,x[2:(length(x)-1)])
  P <- c(0, x,0)
  P <- P/sum(P) #Make sure it sums to 1
  return(P)
}
#proportion fx
prop <- function(data) {
  x = table(unlist(data[,2]))
  prop_1 <- x[1] / (x[1]+x[2])
  return(prop_1)
}
ray_thresh <- rayDISC(phy = td$phy, thresh_dat, rate.mat = four, root.p = root.obs(prop = prop(thresh_dat), 4))

ray_ARD <- rayDISC(phy=td$phy, thresh_dat, rate)

#ray_thresh$phy$edge.length <- unlist(ray_thresh$phy$edge.length)
#paintBranches(ray_thresh$phy, ray_thresh$phy$edge.length, state = ray_thresh$states)

plotRECON(ray_thresh$phy, ray_thresh$states, piecolors = 
c("yellow", "cornflowerblue", "orange", "dark blue"), cex = 0.40, pie.cex = 0.25, height = 50, width = 12, 
show.tip.label = TRUE, label.offset = 0.4, title = "threshold molar",file = "threshEX.pdf", type = "fan")

#pure discrete plot
#edgey <- unlist(corr_two$phy$edge.length)
#paintBranches(corr_two$phy, edge = corr_two$phy$edge.length, state = corr_two$states)

corr_two_plot <- plotRECON(corr_two$phy, corr_two$states, piecolors = 
            c( "cornflowerblue", "orange"), cex = 0.40, pie.cex = 0.25, height = 50, width = 12, 
          show.tip.label = TRUE, label.offset = 0.4, title = "discrete molar",file = "discreteEX.pdf", type = "fan")

corr_four_plot <- plotRECON(corr_d$phy, corr_d$states, piecolors = 
                             c("yellow", "cornflowerblue", "orange", "dark blue"), cex = 0.40, pie.cex = 0.25, height = 50, width = 12, 
                           show.tip.label = TRUE, label.offset = 0.4, title = "corr 4-state molar",file = "corrEX.pdf", type = "fan")


# mcmc threshold 
library(MCMCglmmRAM)
#making dataset ()
mcmc_dat <- subset()
prior <- 
mcmc.thresh <- MCMCglmm(cbind() ~ trait -1, burnin =10000, nitt = 2e+05,
                     random = ~corg(trait):animal, rcov=~corg(trait):units, pedigree= td$phy, 
                     reduced = TRUE, data = td_glm,  prior = prior1, pr = TRUE, thin = 100,
                     pl = TRUE, family = c("categorical", "gaussian", "gaussian", "gaussian", "gaussian"))

y_dat <- as.factor(setNames(d_dat[,1], rownames(d_dat)))
dotTree(td$phy, y_dat, colors = c("blue","red"))


```